Comprehensive assessment of the climate-induced water scarcity over continental Chile

### 2023-10-08 article_bckp/article.Rmd
working third paragraph introduction

---
title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.com
    affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Hemera Centro de Observación de la Tierra
    organization: Facultad de Ciencias, Universidad Mayor.
    addressline: La Piramide 5750
    city: Santiago
    state: Huechuraba
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally. Chile has been affected by persistent water scarcity which is impacting the hydrological system and vegetation development. Central Chile it is been the focus of research studies due to the diminishing water supply, nevertheless water deficit is expanded beyond. We analyze earth observation data for 2000-2023 to do a comprehensive assessment of water scarcity in Chile.  For the analysis, continental Chile was divided into five zones from north to south.  We used the time series of MODIS for land cover change (LULC), fraction of vegetation cover (FVC), and to derive the drought index zcNDVI and zcET (standardized anomaly of cumulative NDVI and ET through the growing season); and CHELSA v2.1 dataset to derive the standardized precipitation evapotranspiration index (SPEI). We evaluate the interconnection between the productivity drought index (zcNDVI) and water demand (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal correlation of the drought indices with total water storage (TWS) from GRACE (Gravity Recovery and Climate Experiment). Our results showed that ...

keywords: 
  - keyword1
  - keyword2
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p
bibliography: references.bib
biblio-style: "apalike"
linenumbers: false
numbersections: true
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
  word_document: default
---

@@

# Version

This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
However, the official template gets more frequently updated than the 'rticles'
package. Therefore, please make sure prior to paper submission, that you're 
using the most recent .cls, .tex and .bst files 
(available [here](http://www.mdpi.com/authors/latex)).

@@

The sixth assessment report (AR6) from the working group I of the IPCC [@IPCC2021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes and the evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Chile has been facing a persistent rainfall deficit lasting for more than ten years [@Garreaud2017] which has impacted the hydrological system [@Boisier2018], and consequently the vegetation development [@Zambrano2020].

Precipitation is the primary driver of drought that impacts hydrological regimes and vegetation productivity. Thus, it is commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, it has been argued that this definition does not fully address the ecological dimensions [@Crausbay2017]. @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems". The AR6 [@IPCC2021] state that even if global warming is stabilized at 1.5°-2°C many parts of the world will be impacted by more severe agricultural and ecological drought. Central Chile has suffered from crop productivity failure, highlighting the growing season 2007-2008 and 2008-2009 [@Zambrano2016;@Zambrano2018], which impacted an extensive surface. But, in 2019-2020, the drought intensity reached an extreme condition at North 34°S  not seen -at least- for more than 40 years [@Zambrano2020], affecting forest, grassland, and croplands areas. The prolonged lack of precipitation within Central Chile is producing changes in the ecosystem that should study.

Satellite remote sensing [@West2019;@AghaKouchak2015] is the primary method to evaluate how meteorological drought impacts vegetation dynamics. Since the 90's multiple vegetation drought indices have been derived (VCI,[@Kogan1990]; TCI, [@Kogan1995];zNDVI, [@Peters2002]; VegDri, [@Brown2008]) that have allowed making spatiotemporal analysis. Although we can calculate those indices for any time during the year (depending on satellite revisit), there are relevant during the stage vegetation has more activity, the growing season [@Mishra2015].  Although modeling phenology is a complex task, satellites offer strategies that help to address it [@Younes2021; @Vrieling2018;@Cai2017]. Also, the land cover dynamics product MCD12Q2 from the USGS [@Friedl2019] provides some phenology metrics. Some authors have proposed indices aggregated during the season. @Meroni2017 accumulating the fractional active photosynthetic active radiation(FAPAR) between the start (SOS) and the end of the season (EOS) in the Sahel, calculate the zCFAPAR. @Zambrano2018 used the same approach but with the NDVI (Normalized Difference Vegetation Index), derivating the zcNDVI within Central Chile. Besides, land use land cover  (LULC) change can be driven by drought [@Tran2019;@Akinyemi2021]. To analyze those changes, multiple time-series LULC products exist as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC [@ESA2017]. For Chile, 2014 was made a high-resolution land cover at 30m of spatial resolution [@Zhao2016]. The LULC product with the vegetation drought index can help evaluate the impact of drought on the ecosystem.

>>>>

---
title: "Comprehensive assessment of the climate-induced water scarcity over continental
  Chile"
author:
- name: Francisco Zambrano
  email: francisco.zambrano@umayor.com
  affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
  correspondingauthor: yes
  footnote: 1
- name: Bob Security
  email: bob@example.com
  affiliation: Another University
- name: Cat Memes
  email: cat@example.com
  affiliation: Another University
  footnote: 2
- name: Derek Zoolander
  email: derek@example.com
  affiliation: Some Institute of Technology
  footnote: 2
date: "`r Sys.Date()`"
output:
  word_document:
    number_sections: yes
    toc: yes
  rticles::elsevier_article:
    keep_tex: yes
    citation_package: natbib
abstract: "Human-induced greenhouse gas emissions have increased the frequency and/or
  intensity of some weather and climate extremes globally. Chile has been affected
  by persistent water scarcity which is impacting the hydrological system and vegetation
  development. Central Chile it is been the focus of research studies due to the diminishing
  water supply, nevertheless our results evidence that water deficit is expanded beyond.
  We analyze earth observation data for 2000-2023 to make a comprehensive assessment
  of water scarcity in Chile.  For the analysis, continental Chile was divided into
  five zones big north, little north, central, south, and austral zone.  We used the
  time series of MODIS for land cover change (LULC), fraction of vegetation cover
  (FVC), and to derive the drought index zcNDVI and zcET (standardized anomaly of
  cumulative NDVI and ET through the growing season); and CHELSA v2.1 dataset to derive
  the standardized precipitation evapotranspiration index (SPEI). We evaluate the
  interconnection between the productivity drought index (zcNDVI) and water demand
  (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal correlation of
  the drought indices with total water storage (TWS) from GRACE (Gravity Recovery
  and Climate Experiment). Our LULC results showed a increasing trend of 412 $[km^2\\cdot
  yr^{-1}$] of forest in the south zone, a decresasing trend of 24 $[km^2\\cdot yr^{-1}$]
  \ of cropland in the central zone and an increase of 31 $[km^2\\cdot yr^{-1}$] in
  the south zone, and a diminish of 80 $[km^2\\cdot yr^{-1}$] of barrend land in the
  austral zone. \n"
keywords:
- keyword1
- keyword2
journal: Remote Sensing of Environment
address:
- code: Hemera Centro de Observación de la Tierra
  organization: Facultad de Ciencias, Universidad Mayor.
  addressline: La Piramide 5750
  city: Santiago
  state: Huechuraba
  postcode: 123456
  country: United States
- code: Another University
  organization: Department
  addressline: A street 29
  postcode: 2054 NX
  city: Manchester,
  country: The Netherlands
classoption: preprint, 3p
bibliography: references.bib
biblio-style: apalike
linenumbers: no
numbersections: yes
footnote:
- code: 1
  text: This is the first author footnote.
- code: 2
  text: Another author footnote.
---

@@

@@

The sixth assessment report (AR6) of the IPCC [@IPCC2021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes and the evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Chile has been facing a persistent rainfall deficit lasting for more than ten years [@Garreaud2017] which has impacted the hydrological system [@Boisier2018], and consequently the vegetation development [@Zambrano2020].

Precipitation is the primary driver of drought that impacts hydrological regimes and vegetation productivity. Thus, it is commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, it has been argued that this definition does not fully address the ecological dimensions [@Crausbay2017]. @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems". The AR6 [@IPCC2021] state that even if global warming is stabilized at 1.5°-2°C many parts of the world will be impacted by more severe agricultural and ecological drought. Central Chile has suffered from crop productivity failure, highlighting the growing season 2007-2008 and 2008-2009 [@Zambrano2016;@Zambrano2018], which impacted an extensive surface. But, in 2019-2020, the drought intensity reached an extreme condition at North 34°S  not seen -at least- for more than 40 years [@Zambrano2020], affecting forest, grassland, and croplands areas. The prolonged lack of precipitation within Central Chile is producing changes in the ecosystem that must be studied.

Satellite remote sensing [@West2019;@AghaKouchak2015] is the primary method to evaluate how meteorological drought impacts vegetation dynamics. Since the 90's multiple vegetation drought indices have been derived (VCI,[@Kogan1990]; TCI, [@Kogan1995];zNDVI, [@Peters2002]; VegDri, [@Brown2008]) that have allowed making spatiotemporal analysis. Although we can calculate those indices for any time during the year (depending on satellite revisit), there are relevant during the stage vegetation has more activity, the growing season [@Mishra2015].  Although modeling phenology is a complex task, satellites offer strategies that help to address it [@Younes2021; @Vrieling2018;@Cai2017]. Also, the land cover dynamics product MCD12Q2 from the USGS [@Friedl2019] provides some phenology metrics. Some authors have proposed indices aggregated during the season. @Meroni2017 accumulating the fractional active photosynthetic active radiation(FAPAR) between the start (SOS) and the end of the season (EOS) in the Sahel, calculate the zCFAPAR. @Zambrano2018 used the same approach but with the NDVI (Normalized Difference Vegetation Index), derivating the zcNDVI within Central Chile. Besides, land use land cover  (LULC) change can be driven by drought [@Tran2019;@Akinyemi2021]. To analyze those changes, multiple time-series LULC products exist as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC [@ESA2017]. The LULC product with the vegetation drought index can help evaluate the impact of drought on the ecosystem.

### 2023-10-08 article_bckp/article.tex
working third paragraph introduction

  \begin{abstract}
  Human-induced greenhouse gas emissions have increased the frequency
  and/or intensity of some weather and climate extremes globally. Chile
  has been affected by persistent water scarcity which is impacting the
  hydrological system and vegetation development. Central Chile it is
  been the focus of research studies due to the diminishing water
  supply, nevertheless water deficit is expanded beyond. We analyze
  earth observation data for 2000-2023 to do a comprehensive assessment
  of water scarcity in Chile. For the analysis, continental Chile was
  divided into five zones from north to south. We used the time series
  of MODIS for land cover change (LULC), fraction of vegetation cover
  (FVC), and to derive the drought index zcNDVI and zcET (standardized
  anomaly of cumulative NDVI and ET through the growing season); and
  CHELSA v2.1 dataset to derive the standardized precipitation
  evapotranspiration index (SPEI). We evaluate the interconnection
  between the productivity drought index (zcNDVI) and water demand
  (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal
  correlation of the drought indices with total water storage (TWS) from
  GRACE (Gravity Recovery and Climate Experiment). Our results showed
  that \ldots{}
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

\hypertarget{version}{%
\section{Version}\label{version}}

This Rmd-skeleton uses the mdpi Latex template published 2019/02.
However, the official template gets more frequently updated than the
`rticles' package. Therefore, please make sure prior to paper
submission, that you're using the most recent .cls, .tex and .bst files
(available \href{http://www.mdpi.com/authors/latex}{here}).

@@

The sixth assessment report (AR6) from the working group I of the IPCC
\citep{IPCC2021} indicates that human-induced greenhouse gas emissions
have increased the frequency and/or intensity of some weather and
climate extremes and the evidence has been strengthened since AR5
\citep{IPCC2013}. There is high confidence that the increasing global
warning can expand the land area affected by increasing drought
frequency and severity \citep{IPCCCH112021}. Chile has been facing a
persistent rainfall deficit lasting for more than ten years
\citep{Garreaud2017} which has impacted the hydrological system
\citep{Boisier2018}, and consequently the vegetation development
\citep{Zambrano2020}.

Precipitation is the primary driver of drought that impacts hydrological
regimes and vegetation productivity. Thus, it is commonly classified as
meteorological, hydrological, and agricultural \citep{Wilhite1985}.
Lately, it has been argued that this definition does not fully address
the ecological dimensions \citep{Crausbay2017}. \citet{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''. The AR6 \citep{IPCC2021} state that even
if global warming is stabilized at 1.5°-2°C many parts of the world will
be impacted by more severe agricultural and ecological drought. Central
Chile has suffered from crop productivity failure, highlighting the
growing season 2007-2008 and 2008-2009
\citep{Zambrano2016, Zambrano2018}, which impacted an extensive surface.
But, in 2019-2020, the drought intensity reached an extreme condition at
North 34°S not seen -at least- for more than 40 years
\citep{Zambrano2020}, affecting forest, grassland, and croplands areas.
The prolonged lack of precipitation within Central Chile is producing
changes in the ecosystem that should study.

Satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how meteorological drought impacts vegetation
dynamics. Since the 90's multiple vegetation drought indices have been
derived (VCI,\citep{Kogan1990}; TCI, \citep{Kogan1995};zNDVI,
\citep{Peters2002}; VegDri, \citep{Brown2008}) that have allowed making
spatiotemporal analysis. Although we can calculate those indices for any
time during the year (depending on satellite revisit), there are
relevant during the stage vegetation has more activity, the growing
season \citep{Mishra2015}. Although modeling phenology is a complex
task, satellites offer strategies that help to address it
\citep{Younes2021, Vrieling2018, Cai2017}. Also, the land cover dynamics
product MCD12Q2 from the USGS \citep{Friedl2019} provides some phenology
metrics. Some authors have proposed indices aggregated during the
season. \citet{Meroni2017} accumulating the fractional active
photosynthetic active radiation(FAPAR) between the start (SOS) and the
end of the season (EOS) in the Sahel, calculate the zCFAPAR.
\citet{Zambrano2018} used the same approach but with the NDVI
(Normalized Difference Vegetation Index), derivating the zcNDVI within
Central Chile. Besides, land use land cover (LULC) change can be driven
by drought \citep{Tran2019, Akinyemi2021}. To analyze those changes,
multiple time-series LULC products exist as the MCD12Q1
\citep{Friedl2019} and the ESA CCI-LC \citep{ESA2017}. For Chile, 2014
was made a high-resolution land cover at 30m of spatial resolution
\citep{Zhao2016}. The LULC product with the vegetation drought index can
help evaluate the impact of drought on the ecosystem.

@@

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937146e24b.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709384faf70.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709373919fc5.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709321ad9f19.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709379ea79bd.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093c8ff3c5.pdf}\\
norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709337b6241e.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937f0ed48a.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70934b4bd377.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931e4fd155.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932e8b190a.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093305c1274.pdf}\\
zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932bbec995.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093598f66c9.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709370472832.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093ba3805d.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709341ec566.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933b82960c.pdf}\\
zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70936a87e29b.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935b745f29.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709363a8f17a.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932c57d58c.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935fdd829a.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709328c1f26f.pdf}\\
zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709343d0ce35.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093615c29b7.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933872723d.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709369bca642.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709386fdff0.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931a654519.pdf}\\
\bottomrule
\end{tabular}}
\end{table}

>>>>

  \begin{abstract}
  Human-induced greenhouse gas emissions have increased the frequency
  and/or intensity of some weather and climate extremes globally. Chile
  has been affected by persistent water scarcity which is impacting the
  hydrological system and vegetation development. Central Chile it is
  been the focus of research studies due to the diminishing water
  supply, nevertheless our results evidence that water deficit is
  expanded beyond. We analyze earth observation data for 2000-2023 to
  make a comprehensive assessment of water scarcity in Chile. For the
  analysis, continental Chile was divided into five zones big north,
  little north, central, south, and austral zone. We used the time
  series of MODIS for land cover change (LULC), fraction of vegetation
  cover (FVC), and to derive the drought index zcNDVI and zcET
  (standardized anomaly of cumulative NDVI and ET through the growing
  season); and CHELSA v2.1 dataset to derive the standardized
  precipitation evapotranspiration index (SPEI). We evaluate the
  interconnection between the productivity drought index (zcNDVI) and
  water demand (zcET), supply (SPEIs) and FVC. Finally, we analyzed the
  temporal correlation of the drought indices with total water storage
  (TWS) from GRACE (Gravity Recovery and Climate Experiment). Our LULC
  results showed a increasing trend of 412 \([km^2\cdot yr^{-1}\){]} of
  forest in the south zone, a decresasing trend of 24
  \([km^2\cdot yr^{-1}\){]} of cropland in the central zone and an
  increase of 31 \([km^2\cdot yr^{-1}\){]} in the south zone, and a
  diminish of 80 \([km^2\cdot yr^{-1}\){]} of barrend land in the
  austral zone.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

@@

The sixth assessment report (AR6) of the IPCC \citep{IPCC2021} indicates
that human-induced greenhouse gas emissions have increased the frequency
and/or intensity of some weather and climate extremes and the evidence
has been strengthened since AR5 \citep{IPCC2013}. There is high
confidence that the increasing global warning can expand the land area
affected by increasing drought frequency and severity
\citep{IPCCCH112021}. Chile has been facing a persistent rainfall
deficit lasting for more than ten years \citep{Garreaud2017} which has
impacted the hydrological system \citep{Boisier2018}, and consequently
the vegetation development \citep{Zambrano2020}.

Precipitation is the primary driver of drought that impacts hydrological
regimes and vegetation productivity. Thus, it is commonly classified as
meteorological, hydrological, and agricultural \citep{Wilhite1985}.
Lately, it has been argued that this definition does not fully address
the ecological dimensions \citep{Crausbay2017}. \citet{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''. The AR6 \citep{IPCC2021} state that even
if global warming is stabilized at 1.5°-2°C many parts of the world will
be impacted by more severe agricultural and ecological drought. Central
Chile has suffered from crop productivity failure, highlighting the
growing season 2007-2008 and 2008-2009
\citep{Zambrano2016, Zambrano2018}, which impacted an extensive surface.
But, in 2019-2020, the drought intensity reached an extreme condition at
North 34°S not seen -at least- for more than 40 years
\citep{Zambrano2020}, affecting forest, grassland, and croplands areas.
The prolonged lack of precipitation within Central Chile is producing
changes in the ecosystem that must be studied.

Satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how meteorological drought impacts vegetation
dynamics. Since the 90's multiple vegetation drought indices have been
derived (VCI,\citep{Kogan1990}; TCI, \citep{Kogan1995};zNDVI,
\citep{Peters2002}; VegDri, \citep{Brown2008}) that have allowed making
spatiotemporal analysis. Although we can calculate those indices for any
time during the year (depending on satellite revisit), there are
relevant during the stage vegetation has more activity, the growing
season \citep{Mishra2015}. Although modeling phenology is a complex
task, satellites offer strategies that help to address it
\citep{Younes2021, Vrieling2018, Cai2017}. Also, the land cover dynamics
product MCD12Q2 from the USGS \citep{Friedl2019} provides some phenology
metrics. Some authors have proposed indices aggregated during the
season. \citet{Meroni2017} accumulating the fractional active
photosynthetic active radiation(FAPAR) between the start (SOS) and the
end of the season (EOS) in the Sahel, calculate the zCFAPAR.
\citet{Zambrano2018} used the same approach but with the NDVI
(Normalized Difference Vegetation Index), derivating the zcNDVI within
Central Chile. Besides, land use land cover (LULC) change can be driven
by drought \citep{Tran2019, Akinyemi2021}. To analyze those changes,
multiple time-series LULC products exist as the MCD12Q1
\citep{Friedl2019} and the ESA CCI-LC \citep{ESA2017}. The LULC product
with the vegetation drought index can help evaluate the impact of
drought on the ecosystem.

@@

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
big north & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d410cb31f.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d4ddd8049.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d7c9fbffd.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d1a8bb73a.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d474fdeb5.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d269510e4.pdf}\\
little north & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d5d88be43.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d150a4cbe.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d404fa544.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70dca4c474.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d7e3728ac.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d4a63e6a4.pdf}\\
central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d25a3233e.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d57d04753.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d42f71f69.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d1dec138d.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d1c2543ab.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d8528d9c.pdf}\\
south & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d20c19221.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d2c19997d.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d3e82c025.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d5b7a821e.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d79c7dc48.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d41d72265.pdf}\\
austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d435b8d.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d26854a6b.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d5a94aab.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d53a33e6a.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d14d93133.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70deb41525.pdf}\\
\bottomrule
\end{tabular}}
\end{table}
